---
title: "Case Study 1"
author: "Ekim Buyuk; "
date: "9/4/2018"
output: html_document
---

Objectives:

- determine which factors are associated with birth weight and characterize the relationship between each factor and birth weight
- explore whether fit of a linear regression model is adequate or whether robust regression methods are needed
- estimate the impact of eliminating maternal smoking on birth weight in a population

Interim report (9/7): produce a 5 page (maximum) report using R markdown that clearly describes your process for model selection and validation. Provide clear interpretations of the associations between each predictor and birth weight in layperson’s language, including characterization of uncertainty. Figures may be included as appropriate. Code should be fully reproducible.

#Library of Packages

```{r}
library(dplyr)

```

#Data Import and Cleaning

```{r}
#Read in birth data
o_data <- read.csv("Yr1116Birth.csv", 
                   na.strings=c("9","99", "9999"))
#SEX=1 male, 2 female; male=1 male, 0 female
o_data$male=2-o_data$SEX #binary gender for interpretation

birth_data <- na.omit(o_data)

birth_data$GEST_C=birth_data$GEST
birth_data$BWTG_C=birth_data$BWTG
birth_data$GEST_C[birth_data$GEST_C>50]=NA
birth_data$GEST_C[birth_data$GEST_C<20]=NA
birth_data$BWTG_C[birth_data$BWTG_C<500]=NA
bdclean=na.omit(birth_data)

attach(bdclean)
```

#Factors Associated with Birth Weight

To begin determining which factors are associated with birth weight, we first investigated each variable and made a decision on whether we wanted to treat it as continous or categorical. We then did some preliminary exploratory data analysis using scatter plots, box plots, correlation matrices and histograms. We present our findings on the relationship between birth weight and each of these variables below.

#Numerical Variables: Age, gestation, plurality, parity

```{r}
# Parity
pairs(~BWTG + PARITY + PLUR + GEST + MAGE, data = bdclean)

```

#Categorical Variables: Counties, Race, Hispanic Origin and Cigarettes Smoked

##Average number of Cigarettes Smoked per Day Three Months Prior to Pregnancy
```{r}
bdclean$CIGPN[bdclean$CIGPN == 0] = "Zero Pack"
bdclean$CIGPN[bdclean$CIGPN >0 &bdclean$CIGPN <=20 ]="One pack or less"
bdclean$CIGPN[bdclean$CIGPN >20 &bdclean$CIGPN <=40 ]="One pack to two packs"
bdclean$CIGPN[bdclean$CIGPN >40 &bdclean$CIGPN <=60 ]="Two packs to three packs"
bdclean$CIGPN[bdclean$CIGPN >60 &bdclean$CIGPN <=80 ]="Three packs to four packs"
bdclean$CIGPN[bdclean$CIGPN >80 &bdclean$CIGPN <=98 ]="Four packs to five packs"
bdclean

boxplot(BWTG_C~CIGPN, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day
Three Months Prior to Pregnancy", ylab = "Birthweight in Grams")
```


##Average number of Cigarettes Smoked per Day First Three Months of Pregnancy
```{r}
bdclean$CIGFN[bdclean$CIGFN == 0] = "Zero Pack"
bdclean$CIGFN[bdclean$CIGFN >0 &bdclean$CIGFN <=20 ]="One pack or less"
bdclean$CIGFN[bdclean$CIGFN >20 &bdclean$CIGFN <=40 ]="One pack to two packs"
bdclean$CIGFN[bdclean$CIGFN >40 &bdclean$CIGFN <=60 ]="Two packs to three packs"
bdclean$CIGFN[bdclean$CIGFN >60 &bdclean$CIGFN <=80 ]="Three packs to four packs"
bdclean$CIGFN[bdclean$CIGFN >80 &bdclean$CIGFN <=98 ]="Four packs to five packs"

boxplot(BWTG_C~CIGFN, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day
First Three Months of Pregnancy", ylab = "Birthweight in Grams")
```



##Average number of Cigarettes Smoked per Day Second Three Months of Pregnancy
```{r}
bdclean$CIGSN[bdclean$CIGSN == 0] = "Zero Pack"
bdclean$CIGSN[bdclean$CIGSN >0 &bdclean$CIGSN <=20 ]="One pack or less"
bdclean$CIGSN[bdclean$CIGSN >20 &bdclean$CIGSN <=40 ]="One pack to two packs"
bdclean$CIGSN[bdclean$CIGSN >40 &bdclean$CIGSN <=60 ]="Two packs to three packs"
bdclean$CIGSN[bdclean$CIGSN >60 &bdclean$CIGSN <=80 ]="Three packs to four packs"
bdclean$CIGSN[bdclean$CIGSN >80 &bdclean$CIGSN <=98 ]="Four packs to five packs"


boxplot(BWTG_C~CIGSN, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day
Second Three Months of Pregnancy", ylab = "Birthweight in Grams")
```



##Average number of Cigarettes Smoked per Day Third or Last Trimester

```{r}
bdclean$CIGLN[bdclean$CIGLN == 0] = "Zero Pack"
bdclean$CIGLN[bdclean$CIGLN >0 &bdclean$CIGLN <=20 ]="One pack or less"
bdclean$CIGLN[bdclean$CIGLN >20 &bdclean$CIGLN <=40 ]="One pack to two packs"
bdclean$CIGLN[bdclean$CIGLN >40 &bdclean$CIGLN <=60 ]="Two packs to three packs"
bdclean$CIGLN[bdclean$CIGLN >60 &bdclean$CIGLN <=80 ]="Three packs to four packs"
bdclean$CIGLN[bdclean$CIGLN >80 &bdclean$CIGLN <=98 ]="Four packs to five packs"

boxplot(BWTG_C~CIGLN, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day
Third or last Trimester", ylab = "Birthweight in Grams")
```


##Boxplot: gender of child vs birthweight
```{r}
males = subset(bdclean, male==1)
summary(males$BWTG)
females = subset(bdclean, male==0)
summary(females$BWTG)
boxplot(BWTG~male,col="blue",ylab="Birthweight (grams)",xlab="Gender (female = 0, male = 1)")
```

##Boxplot: race of mother vs birthweight
```{r}
boxplot(BWTG~MRACER, col="red",ylab="Birthweight (grams)", xlab="Race of Mother")
```

##Boxplot: hispanic origin of mother vs birthweight
```{r}
boxplot(BWTG~MHISP, col="green",ylab="Birthweight (grams)", xlab="Hispanic Origin of Mother")
```


#Average Number Of Cigarettes Smoked Per Day Variables

We noticed a strange phenomenon when we looked at the frequency of reported values for average number of cigarettes smoked per day. It appeared that the number reported was higher for multiples of 10, and specifically very high for the number 20. At first, we were convinced there was some reporting error, but investigating further and realizing that most cigarette packs have either 20 or 25 cigarettes in them, we realized that responders were most likely responding with the equivalent of 0.5, 1, 1.5, or 2 packs a day of smoking in their heads. Thinking about it this way, the numbers between the multiples now seemed strange! 

In order to work with this variable, we did not think it made sense to treat it as continuous as we wanted to distinguish between non-smokers and smokers. Thus, we instead treated this as a categorical variable and created the bins as follows: 0 cigarettes smoked / 1-20 cigarettes smoked (<1 pack) / 21-40 cigarettes smoked (More than 1 pack but less than 2 packs) and so on.

```{r}

```

Our findings show that there is a ___ relationship between...

#Linear Regression Model 

Potentially square gestation because it appears to be exhibit a logistic relationship with birth weight.

```{r}

```

#Other Models

```{r}


```

#Impact Of Eliminating Maternal Smoking on Birth Weight in the Population

```{r}


```

