---
title: "Case Study 1"
author: "Ekim Buyuk; "
date: "9/4/2018"
output: html_document
---

Objectives:

- determine which factors are associated with birth weight and characterize the relationship between each factor and birth weight
- explore whether fit of a linear regression model is adequate or whether robust regression methods are needed
- estimate the impact of eliminating maternal smoking on birth weight in a population

Interim report (9/7): produce a 5 page (maximum) report using R markdown that clearly describes your process for model selection and validation. Provide clear interpretations of the associations between each predictor and birth weight in layperson’s language, including characterization of uncertainty. Figures may be included as appropriate. Code should be fully reproducible.

#Data Import and Cleaning

```{r}
#Read in birth data
o_data <- read.csv("Yr1116Birth.csv", 
                   na.strings=c("9","99", "9999"))
#SEX=1 male, 2 female; male=1 male, 0 female
o_data$male=2-o_data$SEX #binary gender for interpretation

birth_data <- na.omit(o_data)

birth_data$GEST_C=birth_data$GEST
birth_data$BWTG_C=birth_data$BWTG
birth_data$GEST_C[birth_data$GEST_C>50]=NA
birth_data$GEST_C[birth_data$GEST_C<20]=NA
birth_data$BWTG_C[birth_data$BWTG_C<500]=NA
bdclean=na.omit(birth_data)
```

#Factors Associated with Birth Weight

```{r}


```

#Linear Regression Model 

```{r}


```

#Other Models

```{r}


```

#Impact Of Eliminating Maternal Smoking on Birth Weight in the Population


```{r}


```

