---
title: "Case Study 1"
author: "Ekim Buyuk; "
date: "9/4/2018"
output: html_document
---

Objectives:

- determine which factors are associated with birth weight and characterize the relationship between each factor and birth weight
- explore whether fit of a linear regression model is adequate or whether robust regression methods are needed
- estimate the impact of eliminating maternal smoking on birth weight in a population

Interim report (9/7): produce a 5 page (maximum) report using R markdown that clearly describes your process for model selection and validation. Provide clear interpretations of the associations between each predictor and birth weight in layperson’s language, including characterization of uncertainty. Figures may be included as appropriate. Code should be fully reproducible.

#Library of Packages

```{r, echo=FALSE}
library(dplyr)
library(magrittr)
```

#Data Import and Cleaning

```{r, echo=FALSE}
#Read in birth data
o_data <- read.csv("Yr1116Birth.csv", 
                   na.strings=c("99", "9999"))
o_data$SEX[o_data$SEX == 9] = NA
#Using SEX=1 male, 2 female; create new variable "male". male=1 means male, 0 means female
o_data$male=2-o_data$SEX #binary gender for interpretation

birth_data <- na.omit(o_data)

birth_data$GEST_C=birth_data$GEST
birth_data$BWTG_C=birth_data$BWTG
birth_data$GEST_C[birth_data$GEST_C>50]=NA
birth_data$GEST_C[birth_data$GEST_C<20]=NA
birth_data$BWTG_C[birth_data$BWTG_C<500]=NA
bdclean=na.omit(birth_data)

attach(bdclean)
```

At the start of our data cleaning, we read in the .csv file with birth weight data from North Carolinian mothers in 2011-2016. In the data, missing values were expressed as "9", "99", or "999", so we replaced these occurrences with NAs. We also accounted for outliers by turning illogical values to NAs. For example, we found it implausible for gestation period to take less than 20 months or more than 50 months, and a birth weight of less than 500 grams would be near impossible. Afterwards, we removed the NA values to create a clean data set. We also made sex into a binary variable called "male," in which "male=1" means the baby was male and "male=0" means the baby was female. This was to make modeling, analysis, and interpretation easier.

#Factors Associated with Birth Weight

To begin determining which factors are associated with birth weight, we first investigated each variable and made a decision on whether we wanted to treat it as continous or categorical. We then did some preliminary exploratory data analysis using scatter plots, box plots, correlation matrices and histograms. We present our findings on the relationship between birth weight and each of these variables below.

#Continous Variables: Age, gestation, plurality, parity

```{r, cache = TRUE}
# Parity
#pairs(~BWTG + PARITY + PLUR + GEST + MAGE, data = bdclean)

```

In order to explore the possible relationships between numerical variables and birth weight, we paired and plotted the values in a matrix. We noticed that the birth weight vs. parity and birth weight vs. plurality plots fanned left, suggesting non-constant variance. We could not notice any clear correlation between birth weight and a mother's age. The strongest correlation we saw was in birth weight vs. gestation, which seems to have a slight curve in its positive trend. This would suggest that we need to use a logistic regression for this variable, or transform the gestation variable by squaring it before performing linear regression.

The plots also show relationships between the variables themselves. We visually examined these plots to look for interaction that we may need to test for. However, there was no clear visual evidence of interaction between variables, so we will have to rely on our regression model to determine possible interactions.

#Categorical Variables: Counties, Race, Hispanic Origin and Cigarettes Smoked

##Average Number Of Cigarettes Smoked Per Day Variables

We noticed a strange phenomenon when we looked at the frequency of reported values for average number of cigarettes smoked per day. It appeared that the number reported was higher for multiples of 10, and specifically very high for the number 20. At first, we were convinced there was some reporting error, but investigating further, we realized that most cigarette packs have either 20 or 25 cigarettes in them. Thus, we realized that responders were most likely responding with the equivalent of one-half, one, one-and-a-half, or two packs a day of smoking in their heads. With this in mind, it would be more logical to look at cigarettes by pack, rather than account for each individual cigarette.

In order to work with this variable, we did not think it made sense to treat it as continuous as we wanted to distinguish between non-smokers and smokers, and between mothers who smoked a little and a lot. Thus, we instead treated this as a categorical variable and created the bins as follows: 0 cigarettes smoked / 1-20 cigarettes smoked (<1 pack) / 21-40 cigarettes smoked (More than 1 pack but less than 2 packs) and so on.

###Average number of Cigarettes Smoked per Day Three Months Prior to Pregnancy
```{r}
bdclean$CIGPN_dichotic[bdclean$CIGPN == 0] = 0
bdclean$CIGPN_dichotic[bdclean$CIGPN > 0] = 1

bdclean$CIGPN_bucket[bdclean$CIGPN == 0] = "0"
bdclean$CIGPN_bucket[bdclean$CIGPN >0 &bdclean$CIGPN <=20 ]="1-20"
bdclean$CIGPN_bucket[bdclean$CIGPN >20 &bdclean$CIGPN <=40 ]="20-40 "
bdclean$CIGPN_bucket[bdclean$CIGPN >40 &bdclean$CIGPN <=60 ]="40-60"
bdclean$CIGPN_bucket[bdclean$CIGPN >60 &bdclean$CIGPN <=80 ]="60-80"
bdclean$CIGPN_bucket[bdclean$CIGPN >80 &bdclean$CIGPN <=98 ]="80-98"

boxplot(BWTG_C~CIGPN_dichotic, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day Three Months Prior to Pregnancy (dichotomized)", ylab = "Birthweight in Grams")

boxplot(BWTG_C~CIGPN_bucket, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day Three Months Prior to Pregnancy (in buckets)", ylab = "Birthweight in Grams")
```


###Average number of Cigarettes Smoked per Day First Three Months of Pregnancy
```{r}
bdclean$CIGFN_dichotic[bdclean$CIGFN == 0] = 0
bdclean$CIGFN_dichotic[bdclean$CIGFN > 0] = 1

bdclean$CIGFN_bucket[bdclean$CIGFN == 0] = "0"
bdclean$CIGFN_bucket[bdclean$CIGFN >0 &bdclean$CIGFN <=20 ]="1-20"
bdclean$CIGFN_bucket[bdclean$CIGFN >20 &bdclean$CIGFN <=40 ]="20-40"
bdclean$CIGFN_bucket[bdclean$CIGFN >40 &bdclean$CIGFN <=60 ]="40-60"
bdclean$CIGFN_bucket[bdclean$CIGFN >60 &bdclean$CIGFN <=80 ]="60-80"
bdclean$CIGFN_bucket[bdclean$CIGFN >80 &bdclean$CIGFN <=98 ]="80-98"

boxplot(BWTG_C~CIGFN_dichotic, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day First Three Months of Pregnancy (dichotomized)", ylab = "Birthweight in Grams")

boxplot(BWTG_C~CIGFN_bucket, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day First Three Months of Pregnancy (in buckets)", ylab = "Birthweight in Grams")
```

###Average number of Cigarettes Smoked per Day Second Three Months of Pregnancy
```{r}
bdclean$CIGSN_dichotic[bdclean$CIGSN == 0] = 0
bdclean$CIGSN_dichotic[bdclean$CIGSN > 0] = 1

bdclean$CIGSN_bucket[bdclean$CIGSN == 0] = "0"
bdclean$CIGSN_bucket[bdclean$CIGSN >0 &bdclean$CIGSN <=20 ]="1-20"
bdclean$CIGSN_bucket[bdclean$CIGSN >20 &bdclean$CIGSN <=40 ]="20-40"
bdclean$CIGSN_bucket[bdclean$CIGSN >40 &bdclean$CIGSN <=60 ]="40-60"
bdclean$CIGSN_bucket[bdclean$CIGSN >60 &bdclean$CIGSN <=80 ]="60-80"
bdclean$CIGSN_bucket[bdclean$CIGSN >80 &bdclean$CIGSN <=98 ]="80-98"

boxplot(BWTG_C~CIGSN_dichotic, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day Second Three Months of Pregnancy (dichotomized)", ylab = "Birthweight in Grams")

boxplot(BWTG_C~CIGSN_bucket, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day Second Three Months of Pregnancy (in bucket)", ylab = "Birthweight in Grams")

```

###Average number of Cigarettes Smoked per Day Third or Last Trimester

```{r}
bdclean$CIGLN_dichotic[bdclean$CIGLN == 0] = 0
bdclean$CIGLN_dichotic[bdclean$CIGLN > 0] = 1

bdclean$CIGLN_bucket[bdclean$CIGLN == 0] = "0"
bdclean$CIGLN_bucket[bdclean$CIGLN >0 &bdclean$CIGLN <=20 ]="1-20"
bdclean$CIGLN_bucket[bdclean$CIGLN >20 &bdclean$CIGLN <=40 ]="20-40"
bdclean$CIGLN_bucket[bdclean$CIGLN >40 &bdclean$CIGLN <=60 ]="40-60"
bdclean$CIGLN_bucket[bdclean$CIGLN >60 &bdclean$CIGLN <=80 ]="60-80"
bdclean$CIGLN_bucket[bdclean$CIGLN >80 &bdclean$CIGLN <=98 ]="80-98"

boxplot(BWTG_C~CIGLN_dichotic, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day Third or last Trimester (dichotomized)", ylab = "Birthweight in Grams")

boxplot(BWTG_C~CIGLN_bucket, data = bdclean, xlab = "Average number of Cigarettes Smoked per Day Third or last Trimester (in bucket)", ylab = "Birthweight in Grams")
```

##Boxplot: Gender of Child vs Birthweight
```{r}
males = subset(bdclean, male==1)
summary(males$BWTG)
females = subset(bdclean, male==0)
summary(females$BWTG)

#Standard Deviation of Birthweights for Males
sd(males$BWTG)
#Standard Deviation of Birthweights for Females
sd(females$BWTG)

#Difference in mean weights based on gender
t.test(males$BWTG, females$BWTG)

boxplot(BWTG~male,col="blue",ylab="Birthweight (grams)",xlab="Gender (female = 0, male = 1)")
```

Taking a look at the relationship of the gender of the child v. the birthweight, it does appear that male babies have slightly higher weights than female babies. The median male baby is appxomately 3,374 grams v. 3,250 grams for a female baby. Furthermore, conducting a simple difference in means test, we do find a statistically significant difference between the mean of male babies (~3,322 grams) and female babies (~3,206 grams). Thus, we do think that we should include this variable in our linear regression model, and it does make sense to treat as a categorical variable (female, male). 

##Boxplot: race of mother vs birthweight
```{r}
boxplot(BWTG~MRACER, col="red",ylab="Birthweight (grams)", xlab="Race of Mother")
```

```{r}
summary(aov(BWTG_C ~ as.factor(MRACER), data = bdclean))
```

From this boxplot it seems like the average birthweight is consistent throughout all the categories of mother's race, except that African mothers on average give birth to slightly lighter babies than mothers of other races. Because of these observations we decided to do an ANOVA test (and it appears from the results that at least one mean is significantly different from the others, which suggests there could be a relationship between birthweight and the race of mother). It will make sense to figure out which race deviates most from others in this aspect(#TODO). Currently we decided to keep race of mother as categorical predictor of birthweight.

##Boxplot: hispanic origin of mother vs birthweight
```{r}
boxplot(BWTG~MHISP, col="green",ylab="Birthweight (grams)", xlab="Hispanic Origin of Mother")

summary(aov(BWTG_C ~ as.factor(MHISP), data = bdclean))
```

Similarly from the box plot average birthweight appears to be consistent across different categories of the mother's hispanic origin. An ANOVA test again suggests that there is at least one mean that is significantly different from the others, which means there could be some relationship with between birthweight and the type of hispanic origin of the mother. Again, it will make sense to figure out which category deviates most from others (#TODO). We decided to use hispanic origin of mother as a categorical variable in our model. 

#Total Pregnancies (Including This One) and Plurality

```{r}

bdclean$PARITY_new[bdclean$PARITY == 1] = "first pregnancy"
bdclean$PARITY_new[bdclean$PARITY > 1]= ">1 pregnancy"
boxplot(BWTG_C~PARITY_new,data=bdclean,col="blue",ylab="Birthweight (grams)",xlab="number of pregnancy")

bdclean$PLUR_new[bdclean$PLUR == 1] = "singleton"
bdclean$PLUR_new[bdclean$PLUR >1]=">1 children"
boxplot(BWTG_C~PLUR_new,data=bdclean,col="green",ylab="Birthweight (grams)",xlab="number of children")
```


##Year of Birth

```{r}
boxplot(BWTG_C~YOB)
bdclean %>% group_by(YOB) %>% summarize(mean_birthweight = mean(BWTG_C))
```

Finally, we took a look at the distributions and means of birthweights across different years. Looking at the means, we found no reason to believe that the year of birth had an impact on the birth weight of the mother, and any significance found through an anova test, could probably be attributed to the large sample size. Thus, we chose not to include this variable as a predictor moving forward. 

#Data cleaning and tranformation

```{r}

```

#Linear Regression Model 

For our linear regression model, we wanted to work with the new transformed variables we had created, exluding YOB, and use backward selection. 

```{r}
#Simple Model with Multiple Categories for Different Amounts of Smoking
Simple_Model_1 <- lm(BWTG_C ~ male + as.factor(CORES) + CIGPN_bucket + CIGFN_bucket + CIGSN_bucket + CIGLN_bucket + GEST_C + PLUR_new + MAGE + MRACER + PARITY_new, data = bdclean) 
  
#Simple Model with Smoking as a Binary Variable                  
Simple_Model_2 <- lm(BWTG_C ~ male + as.factor(CORES) + CIGPN_dichotic + CIGFN_dichotic + CIGSN_dichotic + CIGLN_dichotic + GEST_C + PLUR_new + MAGE + MRACER + PARITY_new, data = bdclean)
```

```{r}
summary(Simple_Model_1)
```

```{r}
summary(Simple_Model_2)
```

```{r}
bdclean$GEST_C_sq = (bdclean$GEST_C)^2

Simple_Model_3 <- lm(BWTG_C ~ male + as.factor(CORES) + CIGPN_dichotic + CIGFN_dichotic + CIGSN_dichotic + CIGLN_dichotic + GEST_C + GEST_C_sq + PLUR_new + MAGE + MRACER + PARITY_new, data = bdclean)

summary(Simple_Model_3)
```

Based on this model, we can see that the varaible CIGPN_dichotic is not significant. 

#Let's Try Some Interaction Variables

```{r}
Interaction_Model <- lm(BWTG_C ~ male + as.factor(CORES) + CIGFN_dichotic + CIGSN_dichotic + CIGLN_dichotic + GEST_C + GEST_C_sq + PLUR_new + MAGE + MRACER + PARITY_new + PARITY_new*MAGE + PLUR_new*MAGE + CIGFN_dichotic*MAGE + CIGSN_dichotic*MAGE + CIGLN_dichotic*MAGE + CIGFN_dichotic*MRACER + CIGSN_dichotic*MRACER + CIGLN_dichotic*MRACER, data = bdclean)

summary(Interaction_Model)
```

Backward Selection based on P-value, order of deletion:
1. CIGLN_dichotic*MRACER // P-value=0.577 
2. CIGSN_dichotic:MAGE // P-value = 0.318992 
3. CIGLN_dichotic: // P-value = 1 0.116130   

Final Model:

Interaction_Model_test <- lm(BWTG_C ~ male + as.factor(CORES) + CIGFN_dichotic + CIGSN_dichotic + GEST_C + GEST_C_sq + PLUR_new + MAGE + MRACER + PARITY_new + PARITY_new*MAGE + PLUR_new*MAGE + CIGFN_dichotic*MAGE + CIGLN_dichotic + CIGLN_dichotic*MAGE + CIGFN_dichotic*MRACER + CIGSN_dichotic*MRACER, data = bdclean)

#Impact Of Eliminating Maternal Smoking on Birth Weight in the Population

```{r}

```

